---
title: "breakouts"
author: "Jacob Rasmussen"
date: "2026-06-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(pROC)
library(xgboost)
library(Matrix)
library(randomForest)
library(ranger)
library(hrbrthemes)
library(ggthemes)
library(ggradar)
library(ggrepel)
```


```{r}
# the big dataset
skaters <- read_csv("../data/skaters_2008_to_2024.csv")
skaters <- skaters |> filter(situation == "all") |> arrange(name, season) |> 
  filter(position != "D")
skaters25 <- read_csv("../data/25-26skaters.csv")
```

```{r}
# the dataset with parameters
breakoutskaters <- skaters |> group_by(name) |>
  filter(n() >= 2) |>
  ungroup()
breakoutskaters <- breakoutskaters |> 
  mutate(
    ppg = I_F_points / games_played
  )
breakoutskaters <- breakoutskaters |> 
  mutate(
    next_ppg = lead(ppg),
    next_points = lead(I_F_points), 
    next_gp = lead(games_played)
  )
breakoutskaters <- breakoutskaters |> 
  mutate(
    ppgchange = next_ppg - ppg,
    ppg_pct_change = ppgchange / ppg
  )
breakoutskaters <- breakoutskaters |> 
  group_by(name) |>
  mutate(
    previous_best_points = cummax(lag(I_F_points, default = 0))
  )
season_gp <- skaters |>
  group_by(season) |>
  summarize(total_games = 82) |>
  mutate(
    total_games = case_when(
      season == 2012 ~ 48, # 2012-13 lockout shortened season
      season == 2019 ~ 71,  # 2019–20 COVID season
      season == 2020 ~ 56,  # 2020–21 COVID season
      TRUE ~ total_games
    )
  )
breakoutskaters <- breakoutskaters |> 
  left_join(season_gp, by = "season") |> 
  mutate(
    gpratio = games_played / total_games,
    nextgpratio = lead(gpratio)
  )

breakoutskaters <- breakoutskaters |> 
  mutate(
    breakout = 
      previous_best_points < 60 &
      I_F_points < 60 &
      ppg_pct_change >= .20 &
      ppgchange >= .15 &
      next_points > previous_best_points &
      next_points > I_F_points &
      next_points >= 40 &
      gpratio >= .6 &
      nextgpratio >= .6
  )
breakoutskaters <- breakoutskaters |> 
  select(season, name, breakout, team, position, ppg, next_ppg, ppgchange, ppg_pct_change, I_F_points, next_points, previous_best_points, games_played, next_gp, everything())
```

```{r}
breakouts <- breakoutskaters |>
  filter(breakout) |>
  filter(season != 2024) |> 
  select(name, breakout, season, team, I_F_points, next_points, ppg, next_ppg, games_played, next_gp, gpratio, nextgpratio)
breakouts_sum <- breakouts |>
  group_by(season) |>
  summarize(breakouts = sum(breakout, na.rm = TRUE))
breakouts_sum

```
```{r}
breakoutskaters <- breakoutskaters |>
  mutate(
    shotpctg = I_F_goals / I_F_shotsOnGoal,
    offensivezoneshift_pctg = I_F_oZoneShiftStarts / I_F_shifts
  )
```


```{r}
breakoutskaters <- breakoutskaters |> 
  group_by(name) |> 
  mutate(
    change_points = I_F_goals - lag(I_F_goals),
    change_icetime = icetime - lag(icetime),
    change_xg = I_F_xGoals - lag(I_F_xGoals),
    change_shots = I_F_shotsOnGoal - lag(I_F_shotsOnGoal),
    change_corsi = onIce_corsiPercentage - lag(onIce_corsiPercentage)
  )
```

```{r}
# Age is not apart of the dataset, so to emulate it, we are adding an "experience" column 
breakoutskaters <- breakoutskaters |> 
  group_by(name) |> 
  mutate(
    experience = row_number()
  ) |> 
  ungroup()
```

```{r}
breakoutskaters <- breakoutskaters |> 
  group_by(name) |> 
  mutate(
    points_growth =
      (I_F_points - lag(I_F_points)) /
      pmax(lag(I_F_points),1),
    xg_growth =
      (I_F_xGoals - lag(I_F_xGoals)) /
      pmax(lag(I_F_xGoals),1),
    icetime_growth =
      (icetime - lag(icetime)) /
      pmax(lag(icetime),1),
    points_2yr =
      I_F_points - lag(I_F_points,2),
    xg_2yr =
      I_F_xGoals - lag(I_F_xGoals,2),
    icetime_2yr =
      icetime - lag(icetime,2),
    ppg_change =
      ppg - lag(ppg),
    ppg_growth =
      (ppg - lag(ppg)) /
      pmax(lag(ppg),0.01)
  ) |> 
  ungroup()
```

```{r}
# Beginning the model
model_data <- breakoutskaters |> 
  select(
    breakout, season, name, team, position, next_points, ppg, I_F_points, I_F_goals, onIce_corsiPercentage, onIce_fenwickPercentage, I_F_xGoals, I_F_xPlayContinuedInZone, I_F_flurryAdjustedxGoals, I_F_highDangerShots, I_F_highDangerGoals, I_F_highDangerxGoals, I_F_lowDangerxGoals, I_F_lowDangerShots, I_F_lowDangerGoals, I_F_mediumDangerShots, I_F_mediumDangerxGoals, I_F_mediumDangerShots, shotpctg, offensivezoneshift_pctg, OnIce_F_xGoals, icetime, iceTimeRank, previous_best_points, experience, change_points, change_icetime, change_xg, change_shots, change_corsi, points_growth, xg_growth, icetime_growth, points_2yr, xg_2yr, icetime_2yr, ppg_change, ppg_growth
  ) |> 
  filter(!is.na(breakout))
```

```{r}
train <- model_data |> 
  filter(
    season <= 2020
  )

test <- model_data |> 
  filter(
    season > 2020
  )
```

```{r}
model <- glm(breakout ~ ppg + onIce_corsiPercentage + I_F_goals + I_F_points + I_F_xGoals + I_F_highDangerxGoals + shotpctg + offensivezoneshift_pctg + icetime + previous_best_points, data = train, family = "binomial")
summary(model)
```

```{r}
test$pred_prob <- predict(
  model,
  newdata = test,
  type = "response"
)
```

```{r}
roc(
  test$breakout,
  test$pred_prob
)
```
```{r}
test |>
  arrange(desc(pred_prob))
```

```{r}
# Using xgboost to make our model better
features <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points"
)
train_x <- train[, features]
test_x  <- test[, features]
```

```{r}
train_matrix <- data.matrix(train[, features])
test_matrix <- data.matrix(test[, features])

train_y <- train$breakout
test_y <- test$breakout
```

```{r}
dtrain <- xgb.DMatrix(data = train_matrix, label = train_y, missing = NA)

dtest <- xgb.DMatrix(data = test_matrix, label = test_y,missing = NA)
```


```{r}
model_xgb_v1 <- xgboost(
  x = train_matrix,
  y = train_y,
  objective = "binary:logistic",
  eval_metric = "auc",
  nrounds = 200,
  max_depth = 4,
  learning_rate = 0.05,
  subsample = 0.8,
  colsample_bytree = 0.8
)

test$pred_prob_xgb_v1 <- predict(model_xgb_v1, test_matrix)
roc(test_y, test$pred_prob_xgb_v1)
```
```{r}
xgb.importance(model = model_xgb_v1)
```

- Now the model is in a good place for the baseline, and I think any improvements would come from better data instead of better models. To look into this, I'm going to add change over time stats for important stats in the models to see if the model can accurately predict if players who have been improving continuously are more or less likely to break out and hopefully improve the precision and AUC of our already pretty good xgboost model. 



```{r}
# Building a second version of the model
features2 <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points",
  "experience",
  "change_points",
  "change_icetime",
  "change_xg",
  "change_shots",
  "change_corsi"
)
```

```{r}
train2 <- train |>
  select(breakout, all_of(features2))

test2 <- test |>
  select(breakout, all_of(features2))
```

```{r}
train_y <- train2$breakout
test_y  <- test2$breakout

train_x <- train2[, features2]
test_x  <- test2[, features2]
```

```{r}
train_matrix <- data.matrix(train_x)
test_matrix  <- data.matrix(test_x)
```


```{r}
# our second model!
model_xgb_v2 <- xgboost(
  x = train_matrix,
  y = train_y,
  objective = "binary:logistic",
  eval_metric = "auc",
  nrounds = 300,
  max_depth = 4,
  learning_rate = 0.05,
  subsample = 0.8,
  colsample_bytree = 0.8
)
```

```{r}
test$pred_prob_xgb_v2 <- predict(model_xgb_v2, test_matrix)
xgb.importance(model = model_xgb_v2)
roc(test_y, test$pred_prob_xgb_v2)
```

- Before we optimize xgboost even further, I'm going to just try one more type of model for this project, random forest.

```{r}
features_rf <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points",
  "experience",
  "change_points",
  "change_icetime",
  "change_xg",
  "change_shots",
  "change_corsi"
)
```

```{r}
train_rf <- train |>
  select(breakout, all_of(features_rf))

test_rf <- test |>
  select(breakout, all_of(features_rf))
train_rf$breakout <- as.factor(train_rf$breakout)
test_rf$breakout <- as.factor(test_rf$breakout)
```

```{r}
rf_model <- ranger(
  breakout ~ .,
  data = train_rf,
  probability = TRUE,
  num.trees = 500,
  importance = "impurity"
)
```

```{r}
rf_pred <- predict(rf_model, data = test_rf)
test$pred_prob_rf_v1 <- rf_pred$predictions[,2]
roc(test$breakout,
    test$pred_prob_rf_v1)
```

- Now let's build some more variables for what will be our xgboost_v3 and rf_v2.




```{r}
features_v3 <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points",
  "experience",
  "change_points",
  "change_icetime",
  "change_xg",
  "change_shots",
  "change_corsi",
  "points_growth",
  "xg_growth",
  "icetime_growth",
  "points_2yr",
  "xg_2yr",
  "icetime_2yr", 
  "ppg_change",
  "ppg_growth"
)
```

```{r}
train_v3 <- train |>
  select(breakout, all_of(features_v3)) 
test_v3 <- test |>
  select(breakout, all_of(features_v3)) 
train_y <- train_v3$breakout
test_y <- test_v3$breakout
train_x <- train_v3[, features_v3]
test_x <- test_v3[, features_v3]
train_matrix <- data.matrix(train_x)
test_matrix <- data.matrix(test_x)
```

```{r}
model_xgb_v3 <- xgboost(
  x = train_matrix,
  y = train_y,

  objective = "binary:logistic",
  eval_metric = "auc",

  nrounds = 300,
  max_depth = 4,
  learning_rate = 0.05,

  subsample = 0.8,
  colsample_bytree = 0.8
)
test$pred_prob_xgb_v3 <- predict(
  model_xgb_v3,
  test_matrix
)
roc(response = test_y, predictor = test$pred_prob_xgb_v3)
```
```{r}
train_rf <- train |>
  select(breakout, all_of(features_v3))

test_rf <- test |>
  select(breakout, all_of(features_v3))
train_rf$breakout <- as.factor(train_rf$breakout)
test_rf$breakout <- as.factor(test_rf$breakout)
rf_model <- ranger(
  breakout ~ .,
  data = train_rf,
  probability = TRUE,
  num.trees = 500,
  importance = "impurity"
)
rf_pred <- predict(rf_model, data = test_rf)
test$pred_prob_rf_v2 <- rf_pred$predictions[,2]
roc(test_rf$breakout,
    test$pred_prob_rf_v2)
```


- These features did not work to produce a better model in anything except for top-50 precision, so now we will go back to xgboost_v2 and v1 of rf as the models we will eventually use to make our predictions and refine the parameters before we freeze our model.
```{r}
features2 <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points",
  "experience",
  "change_points",
  "change_icetime",
  "change_xg",
  "change_shots",
  "change_corsi"
)
```

```{r}
train2 <- train |>
  select(breakout, all_of(features2))

test2 <- test |>
  select(breakout, all_of(features2))
```

```{r}
train_y <- train2$breakout
test_y  <- test2$breakout

train_x <- train2[, features2]
test_x  <- test2[, features2]
```

```{r}
train_matrix <- data.matrix(train_x)
test_matrix  <- data.matrix(test_x)
```

```{r}
# refining xgboost_v2
top_k <- function(predictions, truth, k){
  data.frame(
    breakout = truth,
    pred = predictions
  ) |>
    arrange(desc(pred)) |>
    slice_head(n = k) |>
    summarize(correct = sum(breakout)) |>
    pull(correct)
}
grid <- expand.grid(
  max_depth = c(3,4,5),
  learning_rate = c(0.03,0.05,0.1),
  nrounds = c(200,300,500)
)
results <- data.frame()
for(i in 1:nrow(grid)) {

  model <- xgboost(
    x = train_matrix,
    y = train_y,

    objective = "binary:logistic",
    eval_metric = "auc",

    max_depth = grid$max_depth[i],
    learning_rate = grid$learning_rate[i],
    nrounds = grid$nrounds[i],

    subsample = 0.8,
    colsample_bytree = 0.8
  )

  pred <- predict(model, test_matrix)

  auc <- roc(test_y, pred)$auc

  results <- rbind(
    results,
    data.frame(
      max_depth = grid$max_depth[i],
      learning_rate = grid$learning_rate[i],
      nrounds = grid$nrounds[i],
      auc = as.numeric(auc)
    )
  )
}
results |>
  arrange(desc(auc))
```

```{r}
model1 <- xgboost(
  x = train_matrix,
  y = train_y,
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = 3,
  learning_rate = 0.03,
  nrounds = 200,
  subsample = 0.8,
  colsample_bytree = 0.8
)

model2 <- xgboost(
  x = train_matrix,
  y = train_y,
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = 4,
  learning_rate = 0.03,
  nrounds = 200,
  subsample = 0.8,
  colsample_bytree = 0.8
)
pred1 <- predict(model1, test_matrix)
pred2 <- predict(model2, test_matrix)

top_k(pred1, test_y, 10)
top_k(pred1, test_y, 30)
top_k(pred1, test_y, 50)

top_k(pred2, test_y, 10)
top_k(pred2, test_y, 30)
top_k(pred2, test_y, 50)
```
- As the 1st model is slightly better at predicting the top 10, we will use that as our finalized xgboost_v4_precision model. 
```{r}
model_xgb_v4_precision <- xgboost(
  x = train_matrix,
  y = train_y,
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = 3,
  learning_rate = 0.03,
  nrounds = 200,
  subsample = 0.8,
  colsample_bytree = 0.8
)

test$pred_prob_xgb_v4_precision <- predict(model_xgb_v4_precision, test_matrix)
xgb.importance(model = model_xgb_v4_precision)
roc(test_y, test$pred_prob_xgb_v4_precision)
```

- I ran the same experiment but with the features from v3, and found a model that has a higher auc and better top-50 precision, sacrificing top-10 precision however.
```{r}
features_v3 <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points",
  "experience",
  "change_points",
  "change_icetime",
  "change_xg",
  "change_shots",
  "change_corsi",
  "points_growth",
  "xg_growth",
  "icetime_growth",
  "points_2yr",
  "xg_2yr",
  "icetime_2yr", 
  "ppg_change",
  "ppg_growth"
)
train_v3 <- train |>
  select(breakout, all_of(features_v3)) 
test_v3 <- test |>
  select(breakout, all_of(features_v3)) 
train_y <- train_v3$breakout
test_y <- test_v3$breakout
train_x <- train_v3[, features_v3]
test_x <- test_v3[, features_v3]
train_matrix <- data.matrix(train_x)
test_matrix <- data.matrix(test_x)
```



```{r}
model_xgb_v4_coverage <- xgboost(
  x = train_matrix,
  y = train_y,

  objective = "binary:logistic",
  eval_metric = "auc",

  nrounds = 300,
  max_depth = 3,
  learning_rate = 0.03,

  subsample = 0.8,
  colsample_bytree = 0.8
)
test$pred_prob_xgb_v4_coverage <- predict(
  model_xgb_v4_coverage,
  test_matrix
)
roc(response = test_y, predictor = test$pred_prob_xgb_v4_coverage)
top_k(test$pred_prob_xgb_v3, test_y, 10)
top_k(test$pred_prob_xgb_v3, test_y, 30)
top_k(test$pred_prob_xgb_v3, test_y, 50)
```

```{r}
# refining rf_v1
features_rf <- features_rf <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points",
  "experience",
  "change_points",
  "change_icetime",
  "change_xg",
  "change_shots",
  "change_corsi"
) 
train_rf$breakout <- as.numeric(as.character(train_rf$breakout))
test_rf$breakout  <- as.numeric(as.character(test_rf$breakout))
train_rf <- train |>
  select(breakout, all_of(features_rf)) |>
  na.omit()
test_rf <- test |>
  select(breakout, all_of(features_rf)) |>
  na.omit()
train_rf <- train_rf |> filter(!is.na(breakout))
test_rf  <- test_rf |> filter(!is.na(breakout))

grid_rf <- expand.grid(
  mtry = c(3, 5, 7, 10),
  min.node.size = c(1, 5, 10),
  num.trees = c(500, 1000)
)
results_rf <- data.frame()
for(i in 1:nrow(grid_rf)) {

  rf <- ranger(
    breakout ~ .,
    data = train_rf,
    probability = TRUE,

    mtry = grid_rf$mtry[i],
    min.node.size = grid_rf$min.node.size[i],
    num.trees = grid_rf$num.trees[i]
  )

  pred <- predict(rf, data = test_rf)$predictions[,2]

  results_rf <- rbind(
    results_rf,
    data.frame(
      mtry = grid_rf$mtry[i],
      min.node.size = grid_rf$min.node.size[i],
      num.trees = grid_rf$num.trees[i],

      auc = as.numeric(roc(test_rf$breakout, pred)$auc),
      top10 = top_k(pred, test_rf$breakout, 10),
      top30 = top_k(pred, test_rf$breakout, 30),
      top50 = top_k(pred, test_rf$breakout, 50)
    )
  )
}
results_rf |>
  arrange(desc(auc))
```

- As we can see, there was no meaningful improvement on the random forest models, showing that our two xgboost models are the way to go moving forward. Thus, after all of the testing and tuning, we've found our two models for the prediction of the breakout players. 


```{r}
# test df with only the two models we are using
features_v3 <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points",
  "experience",
  "change_points",
  "change_icetime",
  "change_xg",
  "change_shots",
  "change_corsi",
  "points_growth",
  "xg_growth",
  "icetime_growth",
  "points_2yr",
  "xg_2yr",
  "icetime_2yr", 
  "ppg_change",
  "ppg_growth"
)
train_v3 <- train |>
  select(breakout, all_of(features_v3)) 
test_v3 <- test |>
  select(breakout, all_of(features_v3)) 
train_y <- train_v3$breakout
test_y <- test_v3$breakout
train_x <- train_v3[, features_v3]
test_x <- test_v3[, features_v3]
train_matrix <- data.matrix(train_x)
test_matrix <- data.matrix(test_x)
model_xgb_v4_coverage <- xgboost(
  x = train_matrix,
  y = train_y,

  objective = "binary:logistic",
  eval_metric = "auc",

  nrounds = 300,
  max_depth = 3,
  learning_rate = 0.03,

  subsample = 0.8,
  colsample_bytree = 0.8
)
test$pred_prob_xgb_v4_coverage <- predict(
  model_xgb_v4_coverage,
  test_matrix
)
features2 <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points",
  "experience",
  "change_points",
  "change_icetime",
  "change_xg",
  "change_shots",
  "change_corsi"
)
train2 <- train |>
  select(breakout, all_of(features2))

test2 <- test |>
  select(breakout, all_of(features2))
train_y <- train2$breakout
test_y  <- test2$breakout
train_x <- train2[, features2]
test_x  <- test2[, features2]
train_matrix <- data.matrix(train_x)
test_matrix  <- data.matrix(test_x)
model_xgb_v4_precision <- xgboost(
  x = train_matrix,
  y = train_y,
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = 3,
  learning_rate = 0.03,
  nrounds = 200,
  subsample = 0.8,
  colsample_bytree = 0.8
)

test$pred_prob_xgb_v4_precision <- predict(model_xgb_v4_precision, test_matrix)
test$combined_score <- (.6 * test$pred_prob_xgb_v4_precision) + (.4 * test$pred_prob_xgb_v4_coverage)
test <- test |> 
  select(
    breakout, season, pred_prob_xgb_v4_precision, pred_prob_xgb_v4_coverage, combined_score, everything()
  )
```

- Before moving on to the next step, let's take a step back to make some visualizations for our models. 

```{r}
model_results <- data.frame(
  model = c("GLM", "XGBoost v1", "Random Forest v1", "XGBoost v2", "Random Forest v2", "XGBoost v3", "XGBoost v4 Precision", "XGBoost v4 Coverage"),
  auc = c(0.8628, 0.8979, 0.8821, 0.8984, 0.8866, 0.8943, 0.9037, 0.9122),
  top10 = c(2,4,5,4,2,3,6,3),
  top30 = c(7,8,8,8,6,7,10,7),
  top50 = c(10,11,11,12,12,13,12,13)
)
```

```{r}
# AUC graph
ggplot(model_results, aes(x = reorder(model, auc), y = auc)) + geom_col(fill = "black") + coord_flip() +
  labs(
    title = "AUC of Predictive Breakout Models",
    x = "Model",
    y = "AUC",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) + theme_bw() +
  theme(
    plot.caption = element_text(hjust = 0),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

```{r}
ggplot(breakouts_sum) + 
  geom_col(aes(x = season, y = breakouts), fill = "black") +
  scale_x_continuous(breaks = seq(2008,2023, by = 1)) +
  labs(
    title = "Amount of Breakouts By Year",
    subtitle = "There was some fluctuation in amount of breakouts per year",
    x = "Season",
    y = "Number of Breakouts",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) + 
  theme_bw() +
  theme(
    plot.caption = element_text(hjust = 0),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold")
  )
```


```{r}
# Top-k graphs
ggplot(model_results, aes(x = reorder(model, top10), y = top10)) + geom_col(fill = "black") +
  coord_flip() +
  labs(
    title = "Amount of Breakouts in Model's Probability Top 10",
    y = "Breakouts in Top 10",
    x = "Model",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) + theme_fivethirtyeight() +
  theme(
    plot.caption = element_text(hjust = 0),
    plot.title = element_text(hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(
  breaks = seq(0, max(model_results$top30), by = 2)
  )

ggplot(model_results, aes(x = reorder(model, top30), y = top30)) + geom_col(fill = "black") +
  coord_flip() +
  labs(
    title = "Amount of Breakouts in Model's Probability Top 30",
    y = "Breakouts in Top 30",
    x = "Model",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) + theme_fivethirtyeight() +
  theme(
    plot.caption = element_text(hjust = 0),
    plot.title = element_text(hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(
  breaks = seq(0, max(model_results$top30), by = 2)
  )

ggplot(model_results, aes(x = reorder(model, top50), y = top50)) + geom_col(fill = "black") +
  coord_flip() +
  labs(
    title = "Amount of Breakouts in Model's Probability Top 50",
    y = "Breakouts in Top 50",
    x = "Model",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) + theme_fivethirtyeight() +
   theme(
    plot.caption = element_text(hjust = 0),
    plot.title = element_text(hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_y_continuous(
  breaks = seq(0, max(model_results$top30), by = 2)
  )
```

```{r, warning=FALSE}
# trade-off plot
tradeoff <- data.frame(
  model = c("XGBoost Precision", "XGBoost Coverage"),
  top10 = c(6, 3),
  top50 = c(12, 13)
)

tradeoff_long <- tradeoff |>
  pivot_longer(
    cols = c(top10, top50),
    names_to = "metric",
    values_to = "correct_breakouts"
  )

ggplot(tradeoff_long,
       aes(x = model,
           y = correct_breakouts,
           fill = metric)) +
  geom_col(position = "dodge", color = "black") +   
  geom_text(
    aes(label = correct_breakouts),
    position = position_dodge(width = .9),
    vjust = 1.5,
    color = "black",
    fontface = "bold",
    size = 5
  ) +
  scale_fill_manual(
    values = c(
      "top10" = "#1F77B4",
      "top50" = "#FF7F0E"
    ),
    labels = c(
      "top10" = "Top 10",
      "top50" = "Top 50"
    )
  ) +
  labs(
    title = "Precision vs Coverage Comparison",
    x = NULL,
    y = "Breakouts in Top-K",
    fill = NULL,
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) + theme_fivethirtyeight() +
  theme(
    plot.caption = element_text(hjust = 0),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(
    size = 12,
    face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

```{r}
# Predicted Probability Distribution
ggplot(test, aes(x = pred_prob_xgb_v4_precision)) +
  geom_density(
    fill = "#1F77B4",
    alpha = .8,
    color = "black"
  ) + geom_rug(alpha = .3) + 
  labs(
    title = "Distribution of Predicted Breakout Probabilities",
    subtitle = "Most players are assigned a low probability of breaking out",
    x = "Predicted Probability of Breakout",
    y = "Density",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) +
  theme_fivethirtyeight() +
  theme(
    plot.caption = element_text(hjust= 0),
    axis.text.y = element_blank(),
    axis.title.y = element_text(
    size = 12,
    face = "bold"),
    axis.title.x = element_text(
      size = 12,
      face = "bold"
    ),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r}
# Cumulative dist. of breakout probs
ggplot(test, aes(x = pred_prob_xgb_v4_precision)) +
  stat_ecdf(
    linewidth = 1.2,
    color = "black"
  ) +
  theme_stata() +
  scale_y_continuous(
    labels = scales::percent
  ) +
  scale_x_continuous(
    limits = c(0, .4),
    breaks = seq(0, 1, 0.1)
  ) +
  labs(
    title = "Cumulative Distribution of Breakout Probabilities",
    subtitle = "XGBoost v4 Precision model",
    x = "Predicted Breakout Probability",
    y = "Proportion of Players",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) +
  theme(
    plot.caption = element_text(hjust = 0),
    plot.title = element_text( face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 11),
    panel.grid.minor = element_blank()
  )
```

```{r}
# wip edit later
plot_df <- test |>
  arrange(desc(pred_prob_xgb_v4_precision)) |>
  mutate(rank = row_number())

ggplot(plot_df, aes(rank, pred_prob_xgb_v4_precision)) +
  theme_stata() +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Ranked Breakout Probabilities",
    subtitle = "Only a small number of players receive substantial breakout probabilities",
    x = "Player Rank by Predicted Breakout Probability",
    y = "Predicted Probability of Breakout",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) +
  geom_vline(
  xintercept = 30,
  linetype = "dashed",
  colour = "red"
  ) +
  annotate(
  "text",
  x = 30,
  y = 0.32,
  label = "Top 30",
  hjust = -0.1,
  size = 4
) + 
  theme(
    plot.caption = element_text(hjust = 0),
    plot.title = element_text( face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 11),
    panel.grid.minor = element_blank()
  )
```

```{r}
imp <- xgb.importance(model = model_xgb_v4_coverage)
imp_df <- head(imp, 10)
feature_map <- c(
  "I_F_xGoals" = "Expected Goals",
  "previous_best_points" = "Previous Best Points",
  "ppg" = "Points Per Game",
  "icetime" = "Ice Time",
  "onIce_corsiPercentage" = "On-Ice Corsi %",
  "I_F_highDangerxGoals" = "High-Danger xG",
  "points_2yr" = "2 Year Points Δ ",
  "change_xg" = "Δ Expected Goals",
  "offensivezoneshift_pctg" = "Offensive Zone Shift %",
  "ppg_change" = "Δ Points Per Game"
)
imp_df$Feature_clean <- feature_map[imp_df$Feature]
imp_df$Feature_clean <- ifelse(
  is.na(imp_df$Feature_clean),
  imp_df$Feature,
  imp_df$Feature_clean
)
ggplot(imp_df, aes(x = reorder(Feature_clean, Gain), y = Gain)) +
  geom_col(color = "black", fill = "#FF7F0E") +
  coord_flip() +
  theme_bw() +
  labs(
    title = "Top 10 Feature Importance",
    subtitle = "XGBoost v4 Coverage Model",
    x = "Feature",
    y = "Importance (Gain)",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) +
    theme(
    plot.caption = element_text(hjust = 0),
    panel.grid = element_blank(),
    plot.title = element_text(
      face = "bold"
    )
  )

```
```{r}
# jitter plot, wip edit later
ggplot(test, aes(x = pred_prob_xgb_v4_precision, y = breakout)) +
  geom_jitter(
    height = 0.03,
    width = 0.1,
    alpha = 0.20,
    size = 4,
    color = "steelblue"
  ) +
  scale_x_continuous(limits = c(0, 0.4)) +
  scale_y_continuous(breaks = c(0, 1), labels = c("No", "Yes")) +
  labs(
    title = "Predicted Breakout Probability vs Actual Outcome",
    x = "Predicted Probability",
    y = "Breakout?",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details",
    subtitle = "XGBoost v4 Precision Model"
  ) +
  theme_fivethirtyeight() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(hjust = 0),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.y = element_text(
    size = 12,
    face = "bold"),
    axis.title.x = element_text(
      size = 12,
      face = "bold"
    ),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
  )
```


```{r}
test |>
  mutate(bin = cut(pred_prob_xgb_v4_precision,
                   breaks = seq(0, 0.4, by = 0.02))) |>
  group_by(bin) |>
  summarize(
    avg_prob = mean(pred_prob_xgb_v4_precision),
    breakout_rate = mean(breakout, na.rm = TRUE),
    n = n()
  ) |>
  ggplot(aes(x = avg_prob, y = breakout_rate)) +
  geom_line(linewidth = 1.2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  geom_point(size = 2) +
  labs(
    title = "Observed Breakout Rate by Predicted Probability",
    x = "Predicted Probability",
    y = "Actual Breakout Rate",
    caption = "Source: MoneyPuck.com. See Data Documentation Section For More Details"
  ) +
  theme_fivethirtyeight() +
  theme(
    plot.caption =  element_text(hjust = 0),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(
    size = 12,
    face = "bold"),
    axis.title.x = element_text(
      size = 12,
      face = "bold"
    ),
    panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
  )
```



```{r}
# Case Study: Wyatt Johnston
johnston <- test |> 
  filter(name == "Wyatt Johnston")
johnston_current <- skaters25 |>
  filter(name == "Wyatt Johnston") |> 
  filter(situation == "all")
cols_keep <- names(johnston)
johnston_current <- johnston_current |>
  select(any_of(cols_keep))
johnston <- bind_rows(
  johnston,
  johnston_current
)
ggplot(johnston, aes(x = season)) +
  geom_line(
    aes(y = I_F_points,
        color = "Points"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_points,
        color = "Points"),
    size = 3
  ) +
  geom_line(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    size = 3
  ) +
  geom_vline(
    xintercept = 2023,
    linetype = "dashed",
    color = "gray50"
  ) +
  annotate(
    "label",
    x = 2023,
    y = 45,
    label = "Predicted Breakout\nSeason"
  ) +
  scale_color_manual(
  values = c(
    "Points" = "firebrick",
    "Expected Goals" = "steelblue"
  )) +
  scale_x_continuous(breaks = seq(2022,2025, by=1))+
  labs(
    title = "Wyatt Johnston's Rise",
    subtitle = "Both models claimed Wyatt Johnston following his rookie year was the biggest breakout candidate",
    x = "",
    y = "Season Total",
    color = ""
  ) +
  geom_text(
  aes(
    y = I_F_points,
    label = I_F_points,
    color = "Points"
  ),
  nudge_y = 4,
  show.legend = FALSE
  ) +
  geom_text(
  aes(
    y = I_F_xGoals,
    label = round(I_F_xGoals,1),
    color = "Expected Goals"
  ),
  nudge_y = -4,
  show.legend = FALSE
  )  +
  theme_stata() +
  theme(
  panel.grid.major.x = element_blank(),
  panel.grid.minor = element_blank(),
  plot.title = element_text(
    hjust = .5,
    face = "bold",
    size = 16
  ),
  plot.subtitle = element_text(
    hjust = .5,
    size = 11
  )
  )
```

```{r}
scatter_data <- test |>
  select(name,
         season,
         I_F_points,
         I_F_xGoals,
         I_F_highDangerxGoals,
         onIce_corsiPercentage,
         icetime) |>
  pivot_longer(
    cols = c(I_F_xGoals,
             I_F_highDangerxGoals,
             onIce_corsiPercentage,
             icetime),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(metric = recode(metric,
    I_F_xGoals = "Expected Goals",
    I_F_highDangerxGoals = "High Danger xGoals",
    onIce_corsiPercentage = "Corsi%",
    icetime = "Ice Time"
  ))

ggplot(scatter_data,
       aes(x = value,
           y = I_F_points)) +
  geom_point(alpha = .15,
             color = "gray50") +
  geom_point(
    data = filter(scatter_data,
                  name  == "Wyatt Johnston",
                  season == 2022),
    color = "red",
    size = 4
  ) +
  geom_text(
    data = filter(scatter_data,
                  name  == "Wyatt Johnston",
                  season == 2022),
    aes(label = "Wyatt Johnston\n2022-23"),
    color = "red"
  ) +
  labs(
    title = "Wyatt Johnston Compared to the League",
    subtitle = "As a rookie, Johnston was well above average on key metrics",
    x = "",
    y = ""
  ) +
  facet_wrap(~metric,
             scales = "free_x") +
  theme_stata() +
  theme(
    plot.title = element_text(
    hjust = .5,
    face = "bold",
    size = 16
  ),
  )
```

```{r}
ggplot(test, aes(I_F_xGoals)) +
  geom_density(fill = "steelblue") +
  geom_vline(
    xintercept = johnston |>
      filter(season == 2022) |>
      pull(I_F_xGoals),
    color = "green2",
    linewidth = 1.5
  ) +
  annotate(
    "text",
    x = (johnston |>
      filter(season == 2022) |>
      pull(I_F_xGoals)) - .3,
    y = .02, 
    label = "Wyatt Johnston (2022-23)",
    color = "purple"
  ) +
  labs(
    x = "Expected Goals", 
    y = "Density",
    title = "Wyatt Johnston vs the NHL: Expected Goals",
    subtitle = "Johnston had above average xGoals in his rookie season"
  ) +
  theme_fivethirtyeight() + 
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(
    size = 12,
    face = "bold"),
    axis.title.x = element_text(
      size = 12,
      face = "bold"
    ),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  )

ggplot(test, aes(onIce_corsiPercentage)) +
  geom_density(fill = "gray80") +
  geom_vline(
    xintercept = johnston |>
      filter(season == 2022) |>
      pull(onIce_corsiPercentage),
    color = "green2",
    linewidth = 1.5
  ) +
  annotate(
    "text",
    x = (johnston |>
      filter(season == 2022) |>
      pull(onIce_corsiPercentage)) - .03,
    y = 2, 
    label = "Wyatt Johnston (2022-23)",
    color = "purple"
  ) +
  labs(
    x = "Corsi Percentage", 
    y = "Density",
    title = "Wyatt Johnston vs the NHL: Corsi%",
    subtitle = "Johnston had above average Corsi% in his rookie season"
  ) +
  theme_fivethirtyeight() + 
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(
    size = 12,
    face = "bold"),
    axis.title.x = element_text(
      size = 12,
      face = "bold"
    ),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  )



ggplot(test, aes(I_F_highDangerxGoals)) +
  geom_density(fill = "steelblue") +
  geom_vline(
    xintercept = johnston |>
      filter(season == 2022) |>
      pull(I_F_highDangerxGoals),
    color = "green2",
    linewidth = 1.5
  ) +
  annotate(
    "text",
    x = (johnston |>
      filter(season == 2022) |>
      pull(I_F_highDangerxGoals)) - .3,
    y = .05, 
    label = "Wyatt Johnston (2022-23)",
    color = "black"
  ) +
  labs(
    x = "High Danger xGoals", 
    y = "Density",
    title = "Wyatt Johnston vs the NHL: High Danger Expected Goals",
    subtitle = "Johnston had above average High Danger xGoals in his rookie season"
  ) +
  theme_fivethirtyeight() + 
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(
    size = 12,
    face = "bold"),
    axis.title.x = element_text(
      size = 12,
      face = "bold"
    ),
    plot.title = element_text(
      size = 16,
      face = "bold"
    ),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  )
```

- Next player: Pavel Zacha

```{r}
# come back
zacha <- model_data |>
  filter(name == "Pavel Zacha") |>
  left_join(
    test |>
      select(
        name,
        season,
        pred_prob_xgb_v4_precision,
        pred_prob_xgb_v4_coverage
      ),
    by = c("name", "season")
  )
zacha_current <- skaters25 |>
  filter(name == "Pavel Zacha") |> 
  filter(situation == "all")
cols_keep <- names(zacha)
zacha_current <- zacha_current |>
  select(any_of(cols_keep))
zacha <- bind_rows(
  zacha,
  zacha_current
)
```

```{r}
ggplot(zacha, aes(x = season)) +
  geom_line(
    aes(y = I_F_points,
        color = "Points"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_points,
        color = "Points"),
    size = 3
  ) +
  geom_line(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    size = 3
  ) +
  geom_vline(
    xintercept = 2022,
    linetype = "dashed",
    color = "gray50"
  ) +
  geom_point(
  data = filter(zacha,
                season >= 2022),
  aes(y = I_F_points),
  size = 4,
  shape = 21,
  fill = "gold",
  color = "black"
  ) +
  annotate(
  "label",
  x = 2022.15,
  y = 32,
  label = "Model Predicted\nBreakout Season",
  fill = "white",
  color = "gray30"
  ) +
  scale_color_manual(
  values = c(
    "Points" = "firebrick",
    "Expected Goals" = "steelblue"
  )) +
  scale_x_continuous(breaks = seq(2015,2025, by=1))+
  labs(
    title = "Pavel Zacha's Rise",
    subtitle = "Despite middling production, the model trusted Zacha's underlying metrics and correctly predicted his breakout.",
    x = "",
    y = "Season Total",
    color = ""
  ) +
  geom_text(
  aes(
    y = I_F_points,
    label = I_F_points,
    color = "Points"
  ),
  nudge_y = 4.5,
  show.legend = FALSE
  ) +
  geom_text(
  aes(
    y = I_F_xGoals,
    label = round(I_F_xGoals,1),
    color = "Expected Goals"
  ),
  nudge_y = -4,
  show.legend = FALSE
  ) +
  theme_stata() +
  theme(
  panel.grid.major.x = element_blank(),
  panel.grid.minor = element_blank(),
  plot.title = element_text(
    hjust = .5,
    face = "bold",
    size = 16
  ),
  plot.subtitle = element_text(
    hjust = .5,
    size = 9.5
  )
  )
```

```{r}
ggplot(scatter_data,
       aes(x = value,
           y = I_F_points)) +
  geom_point(alpha = .15,
             color = "gray50") +
  geom_point(
    data = filter(scatter_data,
                  name  == "Pavel Zacha",
                  season == 2021),
    color = "red",
    size = 4
  ) +
  geom_text(
    data = filter(scatter_data,
                  name  == "Pavel Zacha",
                  season == 2021),
    aes(label = "Pavel Zacha\n2021-22"),
    color = "red"
  ) +
  labs(
    title = "Pavel Zacha Compared to the League",
    subtitle = "Zacha's underlying metrics showed a player with more upside than he currently was producing.",
    x = "",
    y = "Points"
  ) +
  facet_wrap(~metric,
             scales = "free_x") +
  theme_stata() +
  theme(
  plot.title = element_text(
    hjust = .5,
    face = "bold",
    size = 16
  ),
  plot.subtitle = element_text(
    hjust = .5,
    size = 11
  )
  )
```

```{r}
placeholder <- skaters25 |>
  filter(situation == "all")
cols_keep <- names(model_data)
placeholder <- placeholder |>
  select(any_of(cols_keep))
model_data_combined <- bind_rows(
  model_data,
  placeholder
)
model_data_combined <- model_data_combined |>
  mutate(
    xg_pctile = 100*percent_rank(I_F_xGoals),
    hdxg_pctile = 100*percent_rank(I_F_highDangerxGoals),
    corsi_pctile = 100*percent_rank(onIce_corsiPercentage),
    points_pctile = 100*percent_rank(I_F_points),
    ozone_pctile = 100*percent_rank(offensivezoneshift_pctg),
    ppg_pctile = 100*percent_rank(ppg)
  )
zacha_pct <- model_data_combined |>
  filter(name == "Pavel Zacha", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
zacha_pct$metric <- recode(
  zacha_pct$metric,
  xg_pctile = "Expected Goals",
  hdxg_pctile = "High-Danger xG",
  corsi_pctile = "Corsi %",
  points_pctile = "Points"
)
ggplot(zacha_pct,
       aes(x = season,
           y = percentile,
           color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(
    limits = c(0,100)
  ) +
  scale_x_continuous(breaks = seq(2016,2025, by = 1)) +
  scale_color_manual(
  values = c(
    "Points" = "firebrick",
    "Expected Goals" = "steelblue",
    "High-Danger xG" = "darkgreen",
    "Corsi %" = "goldenrod"
    )
  ) +
  labs(
    title = "Pavel Zacha's Underlying Metrics Overtook his Production Before His Breakout",
    subtitle = "Leaguewide percentile rankings highlight the step Zacha took before his breakout season",
    x = "",
    y = "Percentile",
    color = ""
  ) +
  theme_stata() +
  theme(
    plot.title = element_text(face = 'bold', size = 13)
  )
```

```{r}
# use this for a different player :)
rank_data <- model_data |>
  group_by(season) |>
  mutate(
    xg_rank = rank(-I_F_xGoals)
  ) |>
  ungroup()
zacha_rank <- rank_data |>
  filter(name == "Pavel Zacha")
ggplot(zacha_rank,
       aes(x = season,
           y = xg_rank)) +

  geom_line(
    linewidth = 1.5,
    color = "steelblue"
  ) +

  geom_point(
    size = 3,
    color = "steelblue"
  ) +

  scale_y_reverse() +

  labs(
    title = "Pavel Zacha Climbed the League Rankings Before Breaking Out",
    subtitle = "Leaguewide rank in Expected Goals",
    x = "",
    y = "Rank"
  ) +

  theme_stata()
```



- Next Player: Zach Benson

```{r}
benson <- test |> 
  filter(name == "Zach Benson")
benson_current <- skaters25 |>
  filter(name == "Zach Benson") |> 
  filter(situation == "all")
cols_keep <- names(benson)
benson_current <- benson_current |>
  select(any_of(cols_keep))
benson <- bind_rows(
  benson,
  benson_current
)
```

```{r}
ggplot(benson, aes(x = season)) +
  geom_line(
    aes(y = I_F_points,
        color = "Points"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_points,
        color = "Points"),
    size = 3
  ) +
  geom_line(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    size = 3
  ) +
  geom_vline(
    xintercept = 2025,
    linetype = "dashed",
    color = "gray50"
  ) +
  annotate(
    "label",
    x = 2023.08,
    y = 25,
    label = "First Model\nPrediction"
  ) +
  annotate(
    "label",
    x = 2024.08,
    y = 25,
    label = "Disappointing Year!\nModel Predicts Again"
  ) +
  annotate(
    "label",
    x = 2024.92,
    y = 25,
    label = "Benson\nBreaks Out!"
  ) +
  scale_color_manual(
  values = c(
    "Points" = "firebrick",
    "Expected Goals" = "steelblue"
  )) +
  scale_x_continuous(breaks = seq(2023,2025, by=1))+
  labs(
    title = "Zach Benson's Career Trajectory",
    subtitle = "The models stayed course after a disappointing second season, and it worked out.",
    x = "",
    y = "Season Total",
    color = ""
  ) +
  geom_text(
  aes(
    y = I_F_points,
    label = I_F_points,
    color = "Points"
  ),
  nudge_y = 3,
  show.legend = FALSE
  ) +
  geom_text(
  aes(
    y = I_F_xGoals,
    label = round(I_F_xGoals,1),
    color = "Expected Goals"
  ),
  nudge_y = -3,
  show.legend = FALSE
  ) +
  theme_stata() +
  theme(
  panel.grid.major.x = element_blank(),
  panel.grid.minor = element_blank(),
  plot.title = element_text(
    hjust = .5,
    face = "bold",
    size = 16
  ),
  plot.subtitle = element_text(
    hjust = .5,
    size = 11
  )
  )
```

```{r}
model_data_combined <- model_data_combined |>
  mutate(
    xg_pctile = 100*percent_rank(I_F_xGoals),
    hdxg_pctile = 100*percent_rank(I_F_highDangerxGoals),
    corsi_pctile = 100*percent_rank(onIce_corsiPercentage),
    points_pctile = 100*percent_rank(I_F_points)
  )
benson_pct <- model_data_combined |>
  filter(name == "Zach Benson", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
benson_pct$metric <- recode(
  benson_pct$metric,
  xg_pctile = "Expected Goals",
  hdxg_pctile = "High-Danger xG",
  corsi_pctile = "Corsi %",
  points_pctile = "Points"
)
ggplot(benson_pct,
       aes(x = season,
           y = percentile,
           color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(
    limits = c(0,100)
  ) +
  scale_x_continuous(breaks = seq(2023,2025, by = 1)) +
  scale_color_manual(
  values = c(
    "Points" = "firebrick",
    "Expected Goals" = "steelblue",
    "High-Danger xG" = "darkgreen",
    "Corsi %" = "goldenrod"
    )
  ) +
  labs(
    title = "Zach Benson's Production Eventually Caught up to His Metrics",
    subtitle = "Leaguewide percentile rankings show Benson's promise, before the production came.",
    x = "",
    y = "Percentile",
    color = ""
  ) +
  theme_stata() +
  theme(
    plot.title = element_text(face = 'bold')
  )
```

```{r}
benson_gap <- model_data_combined |>
  filter(name == "Zach Benson") |>
  mutate(
    gap = xg_pctile - points_pctile
  )
ggplot(benson_gap,
       aes(x = season,
           y = gap)) +
  geom_line(
    linewidth = 1.5,
    color = "steelblue"
  ) +
  geom_point(
    size = 4,
    color = "steelblue"
  ) +
  geom_text(
    aes(label = round(gap,1)),
    nudge_y = 1.5,
    size = 4
  ) +
  scale_x_continuous(
    breaks = unique(benson_gap$season)
  ) +
  labs(
    title = "Benson's Production Finally Caught Up to the Process",
    subtitle = "Difference between Expected Goals percentile and Points percentile",
    x = "",
    y = "Percentile Gap"
  ) +
  theme_stata() +
  theme(
    plot.title = element_text(
      hjust = .5,
      face = "bold",
      size = 16
    ),
    plot.subtitle = element_text(
      hjust = .5,
      size = 10
    ),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```


- Connor Zary

```{r}
# come back!
zary <- test |> 
  filter(name == "Connor Zary")
zary_current <- skaters25 |>
  filter(name == "Connor Zary") |> 
  filter(situation == "all")
cols_keep <- names(zary)
zary_current <- zary_current |>
  select(any_of(cols_keep))
zary <- bind_rows(
  zary,
  zary_current
)
```

```{r}
highlight_players <- tibble(
  name = c(
    "Pavel Zacha",
    "Wyatt Johnston",
    "Connor Zary"
  ),
  season = c(
    2021,
    2022,
    2024
  ),
  player_type = c(
    "Pavel Zacha\n2021-22",
    "Wyatt Johnston\n2022-23",
    "Connor Zary\n2024-25"
  )
)
highlight_points <- scatter_data |>
  inner_join(
    highlight_players,
    by = c("name", "season")
  )
```

```{r}
zary_pct <- model_data_combined |>
  filter(name == "Connor Zary", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
zary_pct$metric <- recode(
  zary_pct$metric,
  xg_pctile = "Expected Goals",
  hdxg_pctile = "High-Danger xG",
  corsi_pctile = "Corsi %",
  points_pctile = "Points"
)
ggplot(zary_pct,
       aes(x = season,
           y = percentile,
           color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(
    limits = c(0,100)
  ) +
  scale_x_continuous(breaks = seq(2023,2025, by = 1)) +
  scale_color_manual(
  values = c(
    "Points" = "firebrick",
    "Expected Goals" = "steelblue",
    "High-Danger xG" = "darkgreen",
    "Corsi %" = "goldenrod"
    )
  ) +
  labs(
    title = "Connor Zary Still Can't Find the Step Forward",
    subtitle = "Leaguewide percentile rankings show Zary's promise, despite the production lacking.",
    x = "",
    y = "Percentile",
    color = ""
  ) +
  theme_stata() +
  theme(
    plot.title = element_text(face = 'bold')
  )
```

```{r}
ggplot(scatter_data,
       aes(x = value,
           y = I_F_points)) +
  geom_point(alpha = .15,
             color = "gray50") +
  geom_point(
    data = highlight_points,
    aes(color = player_type),
    size = 4
  ) +
  geom_smooth(
    method="lm",
    se=FALSE,
    color="black"
  ) +
  labs(
    title = "Why the Model Believed in Connor Zary",
    subtitle = "Zary had above average metrics, but couldn't find the production.",
    x = "",
    y = "Points",
    color = ""
  ) +
  scale_color_manual(
    values = c(
      "Pavel Zacha\n2021-22" = "firebrick",
      "Wyatt Johnston\n2022-23" = "forestgreen",
      "Connor Zary\n2024-25" = "royalblue"
    ) 
  ) +
  scale_y_continuous(
    breaks = seq(0,150, by = 50)
  ) +
  facet_wrap(~metric,
             scales = "free_x") +
  theme_stata() +
  theme(
  plot.title = element_text(
    hjust = .5,
    face = "bold",
    size = 16
  ),
  plot.subtitle = element_text(
    hjust = .5,
    size = 11),
  axis.text.y = element_text(size = 6),
  legend.position = "top"
  )
```

```{r}
ggplot(test, aes(I_F_xGoals)) +
  geom_density(fill = "gray80") +
  geom_vline(
    xintercept = zary |>
      filter(season == 2024) |>
      pull(I_F_xGoals),
    color = "red",
    linewidth = 1.5
  ) +
  geom_vline(
    xintercept = johnston |>
      filter(season == 2022) |>
      pull(I_F_xGoals),
    color = "green2",
    linewidth = 1.5
  ) +
  annotate(
    "text",
    x = (johnston |>
      filter(season == 2022) |>
      pull(I_F_xGoals)) + 5.5,
    y = .03, 
    label = "Wyatt Johnston\n2022-23",
    color = "green2"
  ) +
  annotate(
    "text",
    x = (zary |>
      filter(season == 2024) |>
      pull(I_F_xGoals)) - 4.5,
    y = .01, 
    label = "Connor Zary\n2024-25",
    color = "red"
  ) +
  geom_vline(
    xintercept = zacha |>
      filter(season == 2021) |>
      pull(I_F_xGoals),
    color = "black",
    linewidth = 1.5
  ) +
  annotate(
    "text",
    x = (zacha |>
      filter(season == 2022) |>
      pull(I_F_xGoals)) - 2.7,
    y = .02, 
    label = "Pavel Zacha\n2021-22",
    color = "black"
  ) +
  labs(
    x = "Expected Goals", 
    y = "Density",
    title = "Connor Zary vs the NHL: Expected Goals",
    subtitle = ""
  ) +
  theme_fivethirtyeight() + 
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(
    size = 12,
    face = "bold"),
    axis.title.x = element_text(
      size = 12,
      face = "bold"
    ),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  )

ggplot(test, aes(onIce_corsiPercentage)) +
  geom_density(fill = "gray80") +
  geom_vline(
    xintercept = zary |>
      filter(season == 2024) |>
      pull(onIce_corsiPercentage),
    color = "red",
    linewidth = 1.5
  ) +
  geom_vline(
    xintercept = johnston |>
      filter(season == 2022) |>
      pull(onIce_corsiPercentage),
    color = "green4",
    linewidth = 1.5
  ) +
  annotate(
    "text",
    x = (johnston |>
      filter(season == 2022) |>
      pull(onIce_corsiPercentage)) - .08,
    y = 1, 
    label = "Wyatt Johnston\n2022-23",
    color = "green4"
  ) +
  annotate(
    "text",
    x = (zary |>
      filter(season == 2024) |>
      pull(onIce_corsiPercentage)) + .08,
    y = 2, 
    label = "Connor Zary\n2024-25\n and Pavel Zacha\n2021-22",
    color = "red"
  ) +
  labs(
    x = "Corsi Percentage", 
    y = "Density",
    title = "Connor Zary vs the NHL: Corsi%",
    subtitle = ""
  ) +
  theme_fivethirtyeight() + 
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(
    size = 12,
    face = "bold"),
    axis.title.x = element_text(
      size = 12,
      face = "bold"
    ),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  )



ggplot(test, aes(I_F_highDangerxGoals)) +
  geom_density(fill = "steelblue") +
  geom_vline(
    xintercept = zary |>
      filter(season == 2024) |>
      pull(I_F_highDangerxGoals),
    color = "red",
    linewidth = 1.5
  ) +
  geom_vline(
    xintercept = johnston |>
      filter(season == 2022) |>
      pull(I_F_highDangerxGoals),
    color = "green4",
    linewidth = 1.5
  ) +
  annotate(
    "text",
    x = (johnston |>
      filter(season == 2022) |>
      pull(I_F_highDangerxGoals)) + 3,
    y = .06, 
    label = "Wyatt Johnston\n2022-23",
    color = "green4"
  ) +
  geom_vline(
    xintercept = zacha |>
      filter(season == 2021) |>
      pull(I_F_highDangerxGoals),
    color = "black",
    linewidth = 1.5
  ) +
  annotate(
    "text",
    x = (zacha |>
      filter(season == 2021) |>
      pull(I_F_highDangerxGoals)) - 2.5,
    y = .02, 
    label = "Pavel Zacha\n2021-22",
    color = "black"
  ) +
  annotate(
    "text",
    x = (zary |>
      filter(season == 2024) |>
      pull(I_F_highDangerxGoals)) - 2.5,
    y = .05, 
    label = "Connor Zary\n2024-25",
    color = "red"
  ) +
  labs(
    x = "High Danger xGoals", 
    y = "Density",
    title = "Connor Zary vs the NHL: High Danger Expected Goals",
    subtitle = ""
  ) +
  theme_fivethirtyeight() + 
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(
    size = 12,
    face = "bold"),
    axis.title.x = element_text(
      size = 12,
      face = "bold"
    ),
    plot.title = element_text(
      size = 16,
      face = "bold"
    ),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  )
```

```{r}
compare_df <- model_data |>
  filter(
    (name == "Wyatt Johnston" & season == 2022) |
    (name == "Connor Zary" & season == 2024)
  ) |>
  select(
    name,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile
  ) |>
  pivot_longer(
    -name,
    names_to = "metric",
    values_to = "percentile"
  )
compare_df$metric <- recode(
  compare_df$metric,
  xg_pctile = "Expected Goals",
  hdxg_pctile = "High-Danger xG",
  corsi_pctile = "Corsi %"
)
ggplot(compare_df,
       aes(
         x = metric,
         y = percentile,
         fill = name
       )) +
  geom_col(
    position = "dodge",
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "Wyatt Johnston" = "forestgreen",
      "Connor Zary" = "royalblue"
    )
  ) +
  labs(
    title = "Zary and Johnston Had Similar Underlying Profiles",
    subtitle = "Great metrics doesn't guarantee production, though.",
    x = "",
    y = "League Percentile",
    fill = ""
  ) +
  theme_fivethirtyeight() +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 16
    ),
    plot.subtitle = element_text(
      size = 11
    ),
    axis.title.y = element_text(
      size = 12,
      face = "bold"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  )
```

- Will Cuylle 
```{r}
cuylle <- test |> 
  filter(name == "Will Cuylle")
cuylle_current <- skaters25 |>
  filter(name == "Will Cuylle") |> 
  filter(situation == "all")
cols_keep <- names(cuylle)
cuylle_current <- cuylle_current |>
  select(any_of(cols_keep))
cuylle <- bind_rows(
  cuylle,
  cuylle_current
)
```

```{r}
ggplot(cuylle, aes(x = season)) +
  geom_line(
    aes(y = I_F_points,
        color = "Points"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_points,
        color = "Points"),
    size = 3
  ) +
  geom_line(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    size = 3
  ) +
  geom_vline(
    xintercept = 2024,
    linetype = "dashed",
    color = "gray50"
  ) +
  annotate(
  "label",
  x = 2024.15,
  y = 32,
  label = "Unanticipated\nBreakout Season",
  fill = "white",
  color = "gray30"
  ) +
  scale_color_manual(
  values = c(
    "Points" = "firebrick",
    "Expected Goals" = "steelblue"
  )) +
  scale_x_continuous(breaks = seq(2015,2025, by=1))+
  labs(
    title = "Model Missed Will Cuylle's Breakout",
    subtitle = "Will Cuylle Became a Productive Player, Which the Model Couldn't Predict",
    x = "",
    y = "Season Total",
    color = ""
  ) +
  geom_text(
  aes(
    y = I_F_points,
    label = I_F_points,
    color = "Points"
  ),
  nudge_y = 4,
  show.legend = FALSE
  ) +
  geom_text(
  aes(
    y = I_F_xGoals,
    label = round(I_F_xGoals,1),
    color = "Expected Goals"
  ),
  nudge_y = -4,
  show.legend = FALSE
  )  +
  theme_stata() +
  theme(
  panel.grid.major.x = element_blank(),
  panel.grid.minor = element_blank(),
  plot.title = element_text(
    hjust = .5,
    face = "bold",
    size = 16
  ),
  plot.subtitle = element_text(
    hjust = .5,
    size = 9.5
  )
  )
```

```{r}
cuylle_pct <- model_data_combined |>
  filter(name == "Will Cuylle") |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
cuylle_pct$metric <- recode(
  cuylle_pct$metric,
  xg_pctile = "Expected Goals",
  hdxg_pctile = "High-Danger xG",
  corsi_pctile = "Corsi %",
  points_pctile = "Points"
)



ggplot(cuylle_pct,
       aes(
         season,
         percentile,
         color = metric
       )) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_vline(
    xintercept = 2024,
    linetype = "dashed",
    color = "gray50"
  ) +
  scale_y_continuous(
    limits = c(0,100)
  ) +
  scale_color_manual(
    values = c(
      "Points" = "firebrick",
      "Expected Goals" = "steelblue",
      "High-Danger xG" = "darkgreen",
      "Corsi %" = "goldenrod"
    )
  )  +
  scale_x_continuous(
    breaks = seq(2022,2025,by=1)
  ) +
  labs(
    title = "Cuylle Showed Meaningful Improvement in Metrics",
    subtitle = "While his Corsi% failed to improve, his other metrics rose him to a breakout season.",
    x = "",
    y = "League Percentile",
    color = "",
    linetype = ""
  ) +
  theme_stata() +
  theme(
    plot.title = element_text(face = 'bold')
  )
```

```{r}
cuylle[4,"shotpctg"] <- .127
ggplot(cuylle, aes(x = iceTimeRank, y = shotpctg)) +
  geom_point(size = 4) +
  geom_path(
  arrow = arrow(length = unit(.15,"inches"))) +
  labs(
    title = "Increased Puck Luck and Opportunity Fueled Cuylle's Breakout",
    subtitle = "A great increase in Shooting% and Ice Time Gave Cuylle the opportunity to shine.",
    x = "Ice Time Ranking Among NHL Players",
    y = "Shooting %"
  ) +
  geom_text_repel(
    aes(label = season),
    size = 3.5,
    box.padding = 0.3,
    point.padding = 0.3,
    max.overlaps = Inf
  ) +
  theme_fivethirtyeight() +
    theme(
    plot.title = element_text(
      face = "bold",
      size = 15
    ),
    plot.subtitle = element_text(
      size = 11
    ),
    axis.title.y = element_text(
      size = 12,
      face = "bold"),
    axis.title.x = element_text(
      size = 12,
      face = "bold"),
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
    )
```



- Eetu Luostarinen
```{r}
luostarinen <- model_data |>
  filter(name == "Eetu Luostarinen") |>
  left_join(
    test |>
      select(
        name,
        season,
        pred_prob_xgb_v4_precision,
        pred_prob_xgb_v4_coverage
      ),
    by = c("name", "season")
  )
luostarinen_current <- skaters25 |>
  filter(name == "Eetu Luostarinen") |> 
  filter(situation == "all")
cols_keep <- names(luostarinen)
luostarinen_current <- luostarinen_current |>
  select(any_of(cols_keep))
luostarinen <- bind_rows(
  luostarinen,
  luostarinen_current
)
```


```{r}
ggplot(luostarinen, aes(x = season)) +
  geom_line(
    aes(y = I_F_points,
        color = "Points"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_points,
        color = "Points"),
    size = 3
  ) +
  geom_line(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    size = 3
  ) +
  geom_vline(
    xintercept = 2022,
    linetype = "dashed",
    color = "gray50"
  ) +
  annotate(
  "label",
  x = 2022.15,
  y = 25,
  label = "Model Missed\nBreakout Season",
  fill = "white",
  color = "gray30"
  ) +
  scale_color_manual(
  values = c(
    "Points" = "firebrick",
    "Expected Goals" = "steelblue"
  )) +
  scale_x_continuous(breaks = seq(2015,2025, by=1))+
  labs(
    title = "Eetu Luostarinen: One-Hit Wonder the Model Missed",
    subtitle = "Despite a high upside 2022 that the model failed to predict, Luostarinen has not repeated production",
    x = "",
    y = "Season Total",
    color = ""
  ) +
  geom_text(
  aes(
    y = I_F_points,
    label = I_F_points,
    color = "Points"
  ),
  nudge_y = 4,
  show.legend = FALSE
  ) +
  geom_text(
  aes(
    y = I_F_xGoals,
    label = round(I_F_xGoals,1),
    color = "Expected Goals"
  ),
  nudge_y = -4,
  show.legend = FALSE
  )  +
  theme_stata() +
  theme(
  panel.grid.major.x = element_blank(),
  panel.grid.minor = element_blank(),
  plot.title = element_text(
    hjust = .5,
    face = "bold",
    size = 16
  ),
  plot.subtitle = element_text(
    hjust = .5,
    size = 9.5
  )
  )
```

```{r}
luo_pct <- model_data_combined |>
  filter(name == "Eetu Luostarinen", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
luo_pct$metric <- recode(
  luo_pct$metric,
  xg_pctile = "Expected Goals",
  hdxg_pctile = "High-Danger xG",
  corsi_pctile = "Corsi %",
  points_pctile = "Points"
)

ggplot(luo_pct,
       aes(
         season,
         percentile,
         color = metric,
         linetype = metric
       )) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_vline(
    xintercept = 2022,
    linetype = "dashed",
    color = "gray50"
  ) +
  scale_y_continuous(
    limits = c(0,100)
  ) +
  scale_color_manual(
    values = c(
      "Points" = "firebrick",
      "Expected Goals" = "steelblue",
      "High-Danger xG" = "darkgreen",
      "Corsi %" = "goldenrod"
    )
  ) +
  scale_linetype_manual(
    values = c(
      "Points" = "solid",
      "Expected Goals" = "solid",
      "High-Danger xG" = "solid",
      "Corsi %" = "solid"
    )
  ) +
  scale_x_continuous(
    breaks = seq(2019,2025,by=1)
  ) +
  labs(
    title = "Luostarinen's Production Faded, But His Metrics Did Not",
    subtitle = "Luostarinen's metrics show promise, but his production has been inconsistent.",
    x = "",
    y = "League Percentile",
    color = "",
    linetype = ""
  ) +
  theme_stata()
```

```{r}
luostarinen[7,"shotpctg"] <- .103 # manually adding the missing data
ggplot(luostarinen,
       aes(season, shotpctg)) +
  geom_line(
    linewidth = 1.5,
    color = "firebrick"
  ) +
  geom_point(
    size = 5,
    color = "firebrick"
  ) +
  geom_text(
    aes(
      label = scales::percent(shotpctg, accuracy = 0.1)
    ),
    nudge_y = .01,
    size = 3.5
  ) +
  scale_x_continuous(
    breaks = seq(2019,2025, by = 1)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1)
  ) +
  geom_point(
    aes(color = season == 2022),
  size = 5
  ) +
  scale_color_manual(
    values = c("FALSE" = "firebrick",
             "TRUE" = "goldenrod"),
              guide = "none"
  ) +
  labs(
    title = "Did Shooting Luck Drive Luostarinen's Breakout?",
    subtitle = "Luostarinen's shooting percentage peaked during his breakout 2022 season.",
    x = "",
    y = "Shooting Percentage"
  ) +
  theme_stata() + 
  theme(
    plot.title = element_text(
      face = 'bold'
    )
  )
```

- RUNNING THE MODEL ON 25-26 !

```{r}
features_coverage <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points",
  "experience",
  "change_points",
  "change_icetime",
  "change_xg",
  "change_shots",
  "change_corsi",
  "points_growth",
  "xg_growth",
  "icetime_growth",
  "points_2yr",
  "xg_2yr",
  "icetime_2yr", 
  "ppg_change",
  "ppg_growth"
)
features_precision <- c(
  "ppg",
  "onIce_corsiPercentage",
  "I_F_goals",
  "I_F_xGoals",
  "I_F_highDangerxGoals",
  "shotpctg",
  "offensivezoneshift_pctg",
  "icetime",
  "previous_best_points",
  "experience",
  "change_points",
  "change_icetime",
  "change_xg",
  "change_shots",
  "change_corsi"
)
```

```{r}
# modifying the data to get everything
skaters25[770, "name"] <- "Aatu Raty"
skaters25 <- skaters25 |>
  filter(situation == "all", position != "D") |> 
  mutate(
    shotpctg = I_F_goals / I_F_shotsOnGoal,
    ppg = I_F_points / games_played,
    offensivezoneshift_pctg = I_F_oZoneShiftStarts / I_F_shifts
  )
model_data_combined <- bind_rows(
  model_data,
  skaters25
)
model_data_combined <- model_data_combined |>
  arrange(name, season)
model_data_combined <- model_data_combined |>
  group_by(name) |>
  mutate(
    previous_best_points =
      lag(cummax(I_F_points), default = 0),
    experience =
      row_number() - 1,
    change_points =
      I_F_points - lag(I_F_points),
    change_icetime =
      icetime - lag(icetime),
    change_xg =
      I_F_xGoals - lag(I_F_xGoals),
    change_shots =
      I_F_shotsOnGoal - lag(I_F_shotsOnGoal),
    change_corsi =
      onIce_corsiPercentage - lag(onIce_corsiPercentage),
    points_growth =
      change_points / lag(I_F_points),
    xg_growth =
      change_xg / lag(I_F_xGoals),
    icetime_growth =
      change_icetime / lag(icetime),
    points_2yr =
      I_F_points - lag(I_F_points, 2),
    xg_2yr =
      I_F_xGoals - lag(I_F_xGoals, 2),
    icetime_2yr =
      icetime - lag(icetime, 2),
    ppg_change =
      ppg - lag(ppg),
    ppg_growth =
      ppg_change / lag(ppg)
  ) |>
  ungroup()
predict_data <- model_data_combined |>
  filter(season == 2025) |> 
  select(
    breakout,
    name,
    season,
    team,
    position,
    ppg,
    onIce_corsiPercentage,
    I_F_goals,
    I_F_xGoals,
    I_F_highDangerxGoals,
    shotpctg,
    offensivezoneshift_pctg,
    icetime,
    previous_best_points,
    experience,
    change_points,
    change_icetime,
    change_xg,
    change_shots,
    change_corsi,
    points_growth,
    xg_growth,
    icetime_growth,
    points_2yr,
    xg_2yr,
    icetime_2yr, 
    ppg_change,
    ppg_growth
    )
```

```{r}
# Coverage
predict_matrix_coverage <- as.matrix(
  predict_data[, features_coverage]
)
predict_data$prob_coverage <-
  predict(
    model_xgb_v4_coverage,
    predict_matrix_coverage
  )
predict_data$prob_coverage <- predict(
  model_xgb_v4_coverage,
  predict_matrix_coverage
)
```

```{r}
# Precision
predict_matrix_precision <- as.matrix(
  predict_data[, features_precision]
)
predict_data$prob_precision <-
  predict(
    model_xgb_v4_precision,
    predict_matrix_precision
  )
predict_data$prob_precision <- predict(
  model_xgb_v4_precision,
  predict_matrix_precision
)
```

```{r}
predict_data <- predict_data |> 
  select(
    name,
    team,
    position,
    prob_precision,
    prob_coverage,
    I_F_xGoals,
    everything()
  )
```

```{r}
mackie_pct <- model_data_combined |>
  filter(name == "Mackie Samoskevich", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
```

```{r}
kindel_pct <- model_data_combined |>
  filter(name == "Ben Kindel", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile,
    ozone_pctile,
    ppg_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
kindel_pct
```


```{r}
heineman_pct <- model_data_combined |>
  filter(name == "Emil Heineman", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
heineman_pct
```

```{r}
stankoven_pct <- model_data_combined |>
  filter(name == "Logan Stankoven", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
stankoven_pct
```
```{r}
wood_pct <- model_data_combined |>
  filter(name == "Matthew Wood", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
wood_pct
```
```{r}
dvorsky_pct <- model_data_combined |>
  filter(name == "Dalibor Dvorsky", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
dvorsky_pct
```

```{r}
oconnor_pct <- model_data_combined |>
  filter(name == "Drew O'Connor", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
oconnor_pct
```

```{r}
perreault_pct <- model_data_combined |>
  filter(name == "Gabe Perreault", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |>
  pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
perreault_pct
```

```{r}
# heineman visualization
comparison <- model_data_combined |> 
  filter((name == "Kirill Marchenko" & season == 2022) | (name == "Emil Heineman" & season == 2025)) |> 
  mutate(
    goalshare_scaled = (I_F_goals / I_F_points) * 100,
    corsi_scaled = onIce_corsiPercentage * 100,
    ppg_scaled = ppg * 100,
    shotpctgscaled = shotpctg * 100
  ) |> 
  select(
    name, ppg_scaled, I_F_goals, goalshare_scaled, I_F_xGoals, corsi_scaled, I_F_highDangerxGoals, shotpctgscaled
  ) |> 
  rename(
    "PPG (Scaled)" = ppg_scaled,
    "G" = I_F_goals,
    "G/P (Scaled)" = goalshare_scaled, 
    "Corsi%" = corsi_scaled,
    "xG" = I_F_xGoals,
    "HDxG" = I_F_highDangerxGoals,
    "Shooting %" = shotpctgscaled
  )
comparison_long <- comparison |>
  pivot_longer(
    cols = -name,
    names_to = "metric",
    values_to = "value"
  )
```

```{r}
ggplot(comparison_long,
       aes(x = metric,
           y = value,
           fill = name)) +
  geom_col(
    position = position_dodge(width = .8),
    width = .7,
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "Emil Heineman" = "#F47D30",
      "Kirill Marchenko" = "#002654"
    )
  ) +
  labs(
    title = "Similar Profiles: Kirill Marchenko and Emil Heineman",
    subtitle = "Emil Heineman's 2025 season closely resembles Marchenko's rookie year.",
    x = "",
    y = "Total",
    fill = ""
  ) +
  theme_stata() +
  theme(
    plot.title = element_text(
      hjust = .5,
      face = "bold",
      size = 16
    ),
    plot.subtitle = element_text(
      hjust = .5,
      size = 10.5
    ),
    axis.text.x = element_text(
      size = 10,
      face = "bold"
    ),
    axis.title = element_text(
      face = "bold"
    ),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r}
# stankoven visualization
stankoven_pct <- model_data_combined |>
  filter(name == "Logan Stankoven", season > 2015) |>
  select(
    season,
    xg_pctile,
    hdxg_pctile,
    corsi_pctile,
    points_pctile
  ) |> 
  rename(
    "xG" = xg_pctile,
    "HDxG" = hdxg_pctile,
    "Corsi%"= corsi_pctile,
    "P" = points_pctile
  ) |>
  filter(season == 2025) |> 
    pivot_longer(
    -season,
    names_to = "metric",
    values_to = "percentile"
  )
ggplot(stankoven_pct,
       aes(x = reorder(metric, percentile),
           y = percentile,
           fill = metric)) +
  geom_col(width = .7, color = "black") +
  geom_text(
    aes(label = paste0(round(percentile), "%")),
    hjust = -0.1,
    size = 4.5
  ) +
  scale_fill_manual(
    values = c(
      "P" = "firebrick",
      "xG" = "steelblue",
      "HDxG" = "darkgreen",
      "Corsi%" = "goldenrod"
    )
  ) +
  scale_y_continuous(
    limits = c(0, 105),
    breaks = seq(0,100,20)
  ) +
  coord_flip() +
  labs(
    title = "Logan Stankoven's Points Production is Lagging Behind His Incredible Metrics",
    subtitle = "League percentile rankings entering the 2026-27 season",
    x = "",
    y = "League Percentile",
    fill = ""
  ) +
  theme_stata() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(
      hjust = .5,
      face = "bold",
      size = 13
    ),
    plot.subtitle = element_text(
      hjust = .5,
      size = 10.5
    )
  )
```

```{r}
# samoskevich
samoskevich <- model_data |>
  filter(name == "Mackie Samoskevich") |>
  left_join(
    test |>
      select(
        name,
        season,
        pred_prob_xgb_v4_precision,
        pred_prob_xgb_v4_coverage
      ),
    by = c("name", "season")
  )
samoskevich_current <- skaters25 |>
  filter(name == "Mackie Samoskevich") |> 
  filter(situation == "all")
cols_keep <- names(samoskevich)
samoskevich_current <- samoskevich_current |>
  select(any_of(cols_keep))
samoskevich <- bind_rows(
  samoskevich,
  samoskevich_current
)
```


```{r}
ggplot(samoskevich, aes(x = season)) +
  geom_line(
    aes(y = I_F_points,
        color = "Points"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_points,
        color = "Points"),
    size = 3
  ) +
  geom_line(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    linewidth = 1.5
  ) +
  geom_point(
    aes(y = I_F_xGoals,
        color = "Expected Goals"),
    size = 3
  ) +
  annotate(
  "label",
  x = 2024.875,
  y = 8,
  label = "Will He\n Breakout?",
  fill = "white",
  color = "gray30"
  ) +
  scale_color_manual(
  values = c(
    "Points" = "firebrick",
    "Expected Goals" = "steelblue"
  )) +
  scale_x_continuous(breaks = seq(2023,2025, by=1))+
  labs(
    title = "Mackie Samoskevich Has the Makings of a Star",
    subtitle = "Samoskevich has failed to break out fully, but a high xG could indicate his future.",
    x = "",
    y = "Season Total",
    color = ""
  ) +
  geom_text(
  aes(
    y = I_F_points,
    label = I_F_points,
    color = "Points"
  ),
  nudge_y = 4,
  show.legend = FALSE
  ) +
  geom_text(
  aes(
    y = I_F_xGoals,
    label = round(I_F_xGoals,1),
    color = "Expected Goals"
  ),
  nudge_y = -4,
  show.legend = FALSE
  )  +
  theme_stata() +
  theme(
  panel.grid.major.x = element_blank(),
  legend.position = "bottom",
  panel.grid.minor = element_blank(),
  plot.title = element_text(
    hjust = .5,
    face = "bold",
    size = 16
  ),
  plot.subtitle = element_text(
    hjust = .5,
    size = 9.5
  )
  )
```

```{r}
woodscatter_data <- model_data_combined |>
  select(name,
         season,
         I_F_points,
         I_F_xGoals,
         I_F_highDangerxGoals,
         onIce_corsiPercentage,
         icetime) |>
  pivot_longer(
    cols = c(I_F_xGoals,
             I_F_highDangerxGoals,
             onIce_corsiPercentage,
             icetime),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(metric = recode(metric,
    I_F_xGoals = "Expected Goals",
    I_F_highDangerxGoals = "High Danger xGoals",
    onIce_corsiPercentage = "Corsi%",
    icetime = "Ice Time"
  ))
```

```{r}
ggplot(woodscatter_data,
       aes(x = value,
           y = I_F_points)) +
  geom_point(alpha = .15,
             color = "steelblue") +
  geom_point(
    data = filter(woodscatter_data,
                  name  == "Matthew Wood",
                  season == 2025),
    color = "gold1",
    size = 4
  ) +
  geom_text(
    data = filter(woodscatter_data,
                  name  == "Matthew Wood",
                  season == 2025),
    aes(label = "Matthew Wood\n2025-26"),
    color = "black"
  ) +
  labs(
    title = "Matthew Wood Compared to the League",
    subtitle = "Wood was above average in several metrics, but lacking ice time.",
    x = "",
    y = "Total Points"
  ) +
  facet_wrap(~metric,
             scales = "free_x") +
  theme_stata() +
  theme(
    plot.title = element_text(
    hjust = .5,
    face = "bold",
    size = 16
  ),
  )
```

```{r}
kindel_features <- tibble(
  feature = c(
    "xG",
    "HDxG",
    "PPG",
    "Corsi%",
    "O-Zone Shift%"
  ),
  importance = c(
    .15,
    .05,
    .12,
    .06,
    .03
  ),
  percentile = c(
    88,
    94,
    63,
    69,
    56
  )
)
kindel_features <- kindel_features |>
  mutate(
    importance_scaled = importance/max(importance)*100
  )
```

```{r}
ggplot(kindel_features,
       aes(x = reorder(feature, importance_scaled))) +
  geom_col(
    aes(y = importance_scaled),
    fill = "steelblue",
    color = "black",
    alpha = .8,
    width = .7
  ) +
  geom_point(
    aes(y = percentile),
    color = "firebrick",
    size = 4
  ) +
  geom_text(
    aes(
      y = percentile + 5,
      label = paste0(percentile,"%")
    ),
    color = "firebrick"
  ) +
  coord_flip() +
  scale_y_continuous(
    limits = c(0,105),
    name = "Relative Importance / Percentile"
  ) +
  labs(
    title = "Why Ben Kindel Is Liked by the Model",
    subtitle = "Blue bars show the model's importance to the metric. Red dots show Kindel's performance in them.",
    x = ""
  ) +
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(
      face="bold",
      size=16
    ),
    plot.subtitle = element_text(
      size = 10
    )
  )
```

